---
title: "Lab 11 - Smoking during pregnancy"
author: "Ben Hardin"
date: "3/28/2023"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

```{r set-seed}
set.seed(1234)
```

```{r load-data}
data(ncbirths)
```

### Exercise 1

There are 1000 cases (births) in our sample. For each birth, there are 13 variables.

1. Father's age (numerical)
2. Mother's age (numerical)
3. Whether the mother was classified as "young" or "mature" (categorical)
4. Length of pregnancy in weeks (numerical)
5. Whether the birth was classified as premature or full-term (categorical)
6. Number of hospital visits during pregnancy (numerical)
7. Whether the mother was married or not married (categorical)
8. Weight gained by the mother during pregnancy (numerical)
9. Weight of the baby at birth (numerical)
10. Whether the baby was classified as low birthweight or not (categorical)
11. Gender of the baby (categorical)
12. Smoking status of the mother (categorical)
13. Whether the mother is white or not white (categorical)

There are a few outliers in our numerical data. In particular, there are a lot of outliers for the babyweight variable, with a few babies with very high birthweights and a lot of babies with very low birthweights.

```{r examine-data}
tibble(ncbirths)
```

```{r check-outliers, warning=F}
ggplot(ncbirths, aes(y = fage))+
  geom_boxplot()

ggplot(ncbirths, aes(y = mage))+
  geom_boxplot()

ggplot(ncbirths, aes(y = weeks))+
  geom_boxplot()

ggplot(ncbirths, aes(y = visits))+
  geom_boxplot()

ggplot(ncbirths, aes(y = gained))+
  geom_boxplot()

ggplot(ncbirths, aes(y = weight))+
  geom_boxplot()
```

### Exercise 2

The average weight of (presumably) white babies is 7.25 pounds.

```{r dataset}
ncbirths_white <- ncbirths %>%
  filter(whitemom == "white")

mean(ncbirths_white$weight, na.rm = T)
```
### Exercise 3

We want to make an inference about what happens in the population, but we only have a sample from the population, so this is a good situation to use an inferential test.

### Exercise 4

First, i did one bootstrapped sample based on our sample of white NC births, which ended up having a mean = 7.28.

```{r 1-samples}
birth_boot1 <- ncbirths_white %>%
  slice_sample(n = 1000, replace = T)

mean(birth_boot1$weight, na.rm = T)
```

Then, I ran 10,000 bootstrapped samples, and plotted the distribution of these sample means against the distribution of sample means under the null hypothesis.

Then, I computed a p-value for our observed mean weight under the null hypothesis. The p-value for this test is p < .001, indicating that fewer than 0.1% of the bootstrapped sample means under the null hypothesis were as extreme as the observed mean of our sample.

From this, we can conclude that the average weight of white babies born in NC in 2004 is most likely truly lower than the average weight of white babies born in the US in 1995. I also calculated a 95% confidence interval for our samples, which tells us we can be 95% confident that the average weight in the population of white babies born in NC is somewhere between 7.14 to 7.36 pounds.

```{r boot}
set.seed(1234)
birth_boot <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "mean")
```

```{r compare-dists}
mean(birth_boot$stat, na.rm = T)

birth_null <- birth_boot %>%
  mutate(stat = stat + 0.179903)

ggplot()+
  geom_histogram(data = birth_null, aes(x = stat), fill = "grey80", color = "black", alpha = 0.5)+
  geom_histogram(data = birth_boot, aes(x = stat), fill = "blue3", color = "black", alpha = 0.5)+
  theme_classic()
```

```{r hypothesis-test}
observed_weight <- ncbirths_white %>%
  specify(response = weight) %>%
  calculate(stat = "mean")

get_p_value(birth_null, obs_stat = observed_weight, direction = "two-sided")

birth_boot %>%
  summarize(lower = quantile(stat, 0.025),
            upper = quantile(stat, 0.975))
```

### Exercise 5

I made side-by-side boxplots showing baby's weight as a function of whether or not the mother is a smoker. What this shows is that children of smokers and non-smokers showed similar median weights at birth, but weights were generally somewhat lower for children of smoking mothers, and there were more outliers for children of non-smoking mothers.

```{r boxes}
ncbirths %>%
  filter(!is.na(habit)) %>%
ggplot(aes(x = habit, y = weight))+
  geom_boxplot()
```

### Exercise 6

I made a filtered dataset to only have those cases when mothers reported if they were smokers.

```{r filter}
ncbirths_habitgiven <- ncbirths %>%
  filter(!is.na(habit))
```

### Exercise 7

While there is descriptively a difference between the mean weight for children of smoking vs. non-smoking mothers, we want to know if this is a statistically significant difference (e.g., likely to be actually different between the two populations). Thus, we are going to see if we can reject the null hypothesis:

H0: M(smokers) = M(non-smokers)

and find evidence for the alternative hypothesis:

H1: M(smokers) =/= M(non-smokers)

```{r means}
ncbirths_habitgiven %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
```

### Exercise 8

Yes, because we want to infer from samples to a population.

### Exercise 9

I started running 10,000 bootstrapped samples for the distribution of sample mean differences, for the difference between the mean birthweight for the sample of children of non-smokers vs. the sample of children of smokers. Then, I generated a distribution of sample mean differences under the null hypothesis (i.e., the distribution of sample mean differences if the true difference between these two populations = 0). Then, I ran an inference test comparing the distribution under the null to our observed mean difference of -.32. The p-value for this test was p = .02, indicating that about 2% of the bootstrapped sample mean differences under the null hypothesis were at least as extreme as the observed mean difference in our sample.

From this, we can conclude that there is a good chance that these two populations are indeed different. Under a NHST framework, we would reject the null hypothesis. 

```{r smoker-boots}
set.seed(1234)
smoker_boot <- ncbirths_habitgiven %>%
  specify(weight ~ habit) %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c("smoker", "nonsmoker"))
```

```{r mean-diffs}
mean(smoker_boot$stat, na.rm = T)

smoker_null <- smoker_boot %>%
  mutate(stat = stat + 0.3170139)

ggplot()+
  geom_histogram(data = smoker_null, aes(x = stat), fill = "grey80", color = "black", alpha = 0.5)+
  geom_histogram(data =smoker_boot, aes(x = stat), fill = "grey30", color = "black", alpha = 0.5)+
  theme_classic()
```

```{r smoke-test}
observed_diff <- ncbirths_habitgiven %>%
  specify(weight ~ habit) %>%
  calculate(stat = "diff in means", order = c("smoker", "nonsmoker"))

get_p_value(smoker_null, obs_stat = observed_diff, direction = "both")
```

### Exercise 10

Now that we have some suggestion that there is a real difference in birthweight between children of smokers vs. non-smokers, let's get an idea of what the actual difference might be. I calculated 95% confidence intervals for the distribution of mean differences we generated earlier, getting 95% CIs of [-.58, -.06]. This means we can be 95% certain that the mean birthweight in the population of children of nonsmokers is between about 1/2 a pound to .06 pounds lighter than the mean birthweight in the population of children of smokers.

```{r confidence}
get_ci(smoker_boot, level = 0.95)
```

### Exercise 11

I determined the age cutoff for "maturity" in this dataset by 1) plotting histograms of age for mature vs. young mothers, and 2) getting a count for each age in both subsamples. Both of these show that the cutoff for maturity is 35 years.

```{r hist}
ggplot(ncbirths, aes(x = mage, fill = mature))+
geom_histogram(binwidth = 1)+
  theme_classic()

ncbirths %>%
  group_by(mature) %>%
  count(mage)
```

### Exercise 12

```{r proportion}
set.seed(1234)
weight_age <- ncbirths %>%
  specify(response = lowbirthweight, explanatory = mature, success = "not low") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in props", order = c("younger mom", "mature mom"))
```

```{r null-prop}
mean(weight_age$stat, na.rm = T)

weight_age_null <- weight_age %>%
  mutate(stat = stat - (mean(weight_age$stat, na.rm = T)))

ggplot()+
  geom_histogram(data = weight_age_null, aes(x = stat), fill = "grey80", color = "black", alpha = 0.5)+
  geom_histogram(data = weight_age, aes(x = stat), fill = "grey30", color = "black", alpha = 0.5)+
  theme_classic()
```

```{r prop-test}
observed_prop <- ncbirths %>%
  specify(response = lowbirthweight, explanatory = mature, success = "not low") %>%
  calculate(stat = "diff in props", order = c("younger mom", "mature mom"))
            
get_p_value(weight_age_null, obs_stat = observed_prop, direction = "both")
```

### Exercise 13

```{r propconfidence}
get_ci(weight_age, level = 0.95)
```